Introduction to genomics and DNA methylation

### 2015-05-04 Rmarkdown/biology_background.Rmd
Remove italics

In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated _5-methylcytosine_ from nucleic acid prepared from cow thymus \citep{HOTCHKISS:1948va}. Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.

The typical site of DNA methylation is at the C5 carbon position of a cytosine base, hence _5-methylcytosine_. Figure \ref{fig:5mC} shows the structure of 5mC.

@@

_DNMT2_, now known as _TRDMT1_, was once thought to be a DNA methyltransferase but was shown to in fact methylate a small RNA and not DNA \citep{Goll:2006kb}. Another protein, _DNMT3L_, is homologous to _DNMT3a_ and _DNMT3b_ but does contain catalytic domain that is necessary for methyltransferase activity. Instead, DNMT3L is thought to stimulate the activity of _DNMT3a_ and _DNMT3b_ \citep{Jurkowska:2011dt}.

@@

1. The direct removal of the methyl group from _5mC_ is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ $\to$ _5hmC_ $\to$ _5fC_ $\to$ _5caC_ reaction. The removal of _5hmC_, _5fC_ or _5caC_ is biochemically 'easier' than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.

One question raised by the third point is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks. This is an active area of research.

>>>>

In 1945, Hotchkiss ultimately proved Johnson and Coghill correct when he isolated 5-methylcytosine from nucleic acid prepared from cow thymus \citep{HOTCHKISS:1948va}. Using paper chromatography, Hotchkiss demonstrated that methylated cytosine existed and was distinct from conventional cytosine and uracil.

The typical site of DNA methylation is at the C5 carbon position of a cytosine base, hence 5-methylcytosine. Figure \ref{fig:5mC} shows the structure of 5mC.

@@

_DNMT2_, now known as _TRDMT1_, was once thought to be a DNA methyltransferase but was shown to in fact methylate a small RNA and not DNA \citep{Goll:2006kb}. Another protein, _DNMT3L_, is homologous to _DNMT3a_ and _DNMT3b_ but does contain catalytic domain that is necessary for methyltransferase activity. Instead, _DNMT3L_ is thought to stimulate the activity of _DNMT3a_ and _DNMT3b_ \citep{Jurkowska:2011dt}.

@@

1. The direct removal of the methyl group from 5mC is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a 5mC $\to$ 5hmC $\to$ 5fC $\to$ 5caC reaction. The removal of 5hmC, 5fC or 5caC is biochemically 'easier' than the removal of 5mC and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.

One question raised by the third point is whether 5hmC, 5fC and 5caC are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks. This is an active area of research.
